# Inference

## Two-Sample Data
Kwy questions to consider:
1) WHen do I perform a two-sample test?
2) What are the assumptions?
3) How do I interpret and present the results of the test?
4) How do I deal with non-normal data?

```{r}
library(tidyverse)
library(rstatix) #converts stats functions to a tidyverse-friendly format
```

### 1. WHen do I perform a two-sample test?
These two-sample Student’s t-test is used when we have two samples of continuous data where we are trying to find out if the samples came from the same parent distribution or not. This essentially boils down to finding out if there is a difference in the means of the two samples.

For example, suppose we now measure the body lengths of male guppies (in mm) collected from two rivers in Trinidad; the Aripo and the Guanapo. The data are stored in the file data/CS1-twosample.csv. We want to test whether the mean body length differs between samples. We form the following null and alternative hypotheses:
$H_0:$ The mean body length does not differ between the two groups $(\mu A = \mu G)$
$H_1:$ The mean body length differs between the two groups $(\mu A \neq \mu G)$

Given we have two samples, we will use a two-sample test. Since we want to know if our data suggest that the true (ie. population) means are different from each other rather than that one mean is specifcailly bigger or smaller than the other, we will use a two-tailed test. Finally, because the sample sizes are large enough, and because we will assume that the parent populations have eqaul variance (we can check this), we will use Student's t-test.

### 2. Summarize and Visualize the data
```{r}
rivers <- read_csv("data/CS1-twosample.csv")
head(rivers)
summary(rivers)
```
The standard summary does not give much information. Here we have more than one group - Aripo and Guanapo - so it might be helpful to get summary statistics per group. One way of doing this is by using the get_summary_stats() function from the rstatix library.
```{r}
#get common summary stats for the length column
rivers %>%
  group_by(river) %>%
  get_summary_stats(type = "common")
```

This gives a bit more information including sample sizes, ranges, means etc. However, numbers may not always give the best insight into the data, so we should also visualize the data:
```{r}
ggplot(rivers,
       aes(x = river, y = length)) +
  geom_boxplot()
```

The box plot does appear to suggest that the two samples have different means, and moreover that the guppies in Guanapo may be smaller than the guppies in Aripo. It isn’t immediately obvious that the two populations don’t have equal variances though (box plots are not quite the right tool for this), so we continue.

### 3, Assumptions
In order to use a Student’s t-test (and for the results to be strictly valid) we have to make three assumptions:

1) The parent distributions from which the samples are taken are both normally distributed (which would lead to the sample data being normally distributed too).
2) Each data point in the samples is independent of the others.
3) The parent distributions should have the same variance.

In this example the first assumption can be ignored as the sample sizes are large enough. If the samples were smaller then we would use histogram, Q-Q plot, and Shapiro-Wilk test as we did in the one-sample data discussion.

The second point we can do nothing about unless we know how the data were collected, so again we ignore it.

The third point regarding equality of variance can be tested using either Bartlett’s test (if the samples are normally distributed) or Levene’s test (if the samples are not normally distributed).

This is where it gets a bit trickier. Although we don’t care if the samples are normally distributed for the t-test to be valid (because the sample size is big enough to compensate), we do need to know if they are normally distributed in order to decide which variance test to use. So we perform Shapiro-Wilk test on both samples separately.

```{r}
#We can use the filter() function to filter the data by river, then we perform the Shapiro-Wilk test on the length measurement. The shapiro.test() function needs the data in a vector format. We get these by using the pull() function.
rivers %>%
  filter(river == "Aripo") %>%
  pull(length) %>%
  shapiro.test()

rivers %>%
  filter(river == "Guanapo") %>%
  pull(length) %>%
  shapiro.test()
```

We can see that whilst the Guanapo data is probably normally distributed (p = 0.1764 > 0.05), the Aripo data is unlikely to be normally distributed (p = 0.02802 < 0.05). Remember that the p-value gives the probability of observing each sample if the parent population is actually normally distributed.

The Shapiro-Wilk test is quite sensitive to sample size. This means that if you have a large sample then even small deviations from normality will cause the sample to fail the test, whereas smaller samples are allowed to pass with much larger deviations. Here the Aripo data has nearly 40 points in it compared with the Guanapo data and so it is much easier for the Aripo sample to fail compared with the Guanapo data.

We can also look at the Q-Q plots for each:
```{r}
# we group the data by river
# then create a panel per river
# containing the Q-Q plot for that river
ggplot(rivers,
       aes(sample = length)) +
  stat_qq() +
  stat_qq_line(colour = "blue") +
  facet_wrap(facets = vars(river))
```
The Q-Q plots show the opposite of what we found with the Shapiro-Wilk tests: the data for Aripo look pretty normally distributed apart from one data point, whereas the assumption of normality for the Guanapo data is less certain.

What to do? Well, you could be conservative and state that you are not confident that the data in either group are normally distributed. That would be a perfectly reasonable conclusion.

I would personally not have issues with stating that the Aripo data are probably normally distributed enough.
